Learning Constraints and Optimization Criteria

### 2015-10-20 Paper.tex
Without context and review processing

\def\year{2015}
\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{etoolbox}
\usepackage{color}
\usepackage{changepage}
\usepackage{csquotes}
\usepackage{tabularx,booktabs,multirow}
\usepackage{longtable}
\usepackage{framed}

@@

\title{Learning Constraints and Optimization Criteria}
\author{
  Samuel Kolb\\
  KULeuven, Leuven, Belgium\\
  samuel.kolb@cs.kuleuven.be
}

@@

\begin{abstract}
This paper presents a system that is able to automatically learn both constraints from examples and optimization criteria from rankings.
Both hard and soft constraints can be learned and constraints are represented as first order logical clauses.
Choosing such clauses as representations enables the use of ILP techniques such as clausal discovery.
In this research, optimization criteria are represented as weighted clauses.
For all examples, the clause learning system could find the essential constraints efficiently.
The learning system for optimization criteria is shown to achieve accurate results and can learn criteria that identify the correct optimal solution even for few examples and noisy rankings.
Finally, a procedure is outlined that allows weighted first order logic clauses to be used for optimization.
\end{abstract}

\section{Introduction}
This research tackles the learning of constraints and optimization criteria.
Non-experts often experience difficulties when modeling a problem using constraints \cite{Wallace:PrinciplesCP}.
The first goal of this research is to automate the step of modeling a problem, in order to, amongst others, make constraint solving more accessible.
In this approach a few examples of what a user considers valid solutions are used by a computer system to find constraints.
These constraints can then be used by an existing constraint solver.

Constraints are represented as logical clauses.
This representation allows for the use of techniques from the field of Inductive Logic Programming (ILP).
Specifically, a new implementation of the Clausal Discovery algorithm \cite{DeRaedt:ClausalDiscovery} is used.
Research on constraint learning is often done independent of ILP.
By exploiting the parallels between both fields, this research follows the same approach as \cite{Lallouet:LearningCP}.

Hard constraints are used to distinguish solutions to a problem.
So called soft constraints can be used to express properties which do not have to be fulfilled by every solution.
Soft constraints are assigned weights that express their importance.
Such weighted soft constraints can be used as optimization criteria.
Optimal solutions are found by maximizing the sum of the weights of soft constraints that are fulfilled by a solution.

The second goal of this research is to learn from user preference and to formulate corresponding optimization criteria.
By using rankings over a few examples, weighted soft constraints can be found.
Both the learning of constraints as well as the learning of optimization criteria aim to bridge the gap between problems and their formal representation.

The use of weighted soft constraints is inspired by the research of \cite{campigotto2011active}, amongst others.
Their work examines attempts to obtain similar results in an interactive context for propositional constraints.
In this research, first-order logical clauses will be used.

The weighted clauses cannot be directly used by a constraint solver.
Therefore, an approach is outlined that describes how these constraints can be used in an existing solver to obtain an optimal solution in practice.

This research has been conducted during the course of my masters thesis~\cite{kolb2015thesis}.
\section{Background}
In this section a brief overview of the important concepts and relevant literature will be provided.

\subsection{Clauses}
First-order logical clauses consist of a disjunction of literals.
Literals are defined as being either an atom or a negated atom.
Clauses are often expressed using a body and a head: $\mathit{head} \leftarrow \mathit{body}$.
All literals containing a negation are grouped in the body and the other literals form the head of the clause.
For example, $\lnot a_1 \lor \lnot a_2 \lor a_3 \lor a_4$ corresponds to the clause $a_3, a_4 \leftarrow a_1, a_2$.

Atoms are logical concepts that can be assigned a truth value.
In this paper, atoms will be restricted to either the symbols $\mathit{true}$ or $\mathit{false}$ or a predicate.
Predicates consist of a predicate symbol and a set of terms.
Terms are either constants or variables and describe objects in a domain.
Predicates express relations between these objects: $\mathit{symbol}(\mathit{term_1}, ... \mathit{term_n})$.
Clauses are universally quantified and will, in this research, only contain variables.

\subsection{Learning constraints}
Constraint learning is a hard problem.
The search space of constraints is typically very large.
Furthermore, there are usually only a few examples to learn from, unlike in typical machine learning problems.

@@

\subsection{Clausal Discovery}
The clausal discovery algorithm proposed by \cite{DeRaedt:ClausalDiscovery} attempts to learn first order logic clauses from positive examples.
Using a refinement operator, the search space is traversed by incrementally generalizing clauses until they cover all examples, starting from the empty clause~($\square$).
The result set consists of the most specific clauses covering all examples.
This research uses the clausal discovery algorithm on a high level but provides a very different implementation than the original clausal discovery engine.

\subsection{Logic Solver}
The logic system IDP \cite{de2013prototype,wittocx2008idp} is used to calculate whether examples satisfy clauses and is used as a theorem prover.
It supports an extension of first order logic ($FO(\cdot)$) and can also be used to generate new solutions.
Background knowledge is given as an IDP theory.

\subsection{Optimization}
The Boolean Satisfiability Problem (SAT) attempts to determine whether a propositional formula can be satisfied.
Formulas can be rewritten into a special form called the conjunctive normal form (CNF), which consists of a conjunction of clauses (disjunctions).
An extension of this problem is the Maximum Satisfiability Problem (MAX-SAT).
In MAX-SAT the amount of satisfied clauses is maximized.
By assigning positive integer weights to clauses and maximizing the sum of the weights of satisfied clauses, one obtains the weighted MAX-SAT problem.
The work by \cite{campigotto2011active} attempts to learn propositional formulas en weights automatically.
Inspired by their results, this research tries to learn first-order logic optimization criteria.
The resulting optimization problem can be seen as an extension of weighted MAX-SAT in which the clauses are in first-order logic.

\section{Problem Statement}
This research has two main objectives: learning constraints and learning optimization criteria.
The system designed to solve these problems attempts to simplify the process of generating formal representations.
Therefore, it should be easy to use and take into account what kind of information a user can provide.

\begin{framed}
  \noindent
  \begin{minipage}{\textwidth}
    \paragraph*{Goal 1}
    Given a set of examples and a limit $t$, the goal of the constraint learning system is to find maximally specific clauses that are satisfied by at least $t$ of the given examples.
  \end{minipage}
\end{framed}

Examples describe specific solutions.
They contain a set of objects and exhaustively describe all relations (predicates) that hold on these objects.
The learned clauses are independent of the specific objects within the examples and contain only variables.

\begin{example}
  Consider sudoku as an example of a problem where hard constraints are to be found.
  Given one or more solved sudokus, the learning system has to identify the rules of the game.
  Since the learned clauses only contain universally quantified variables, for example $4 \times 4$ sudokus could be used to find constraints.
  The learned constraints are independent of the size of the sudoku and can also be used to solve $9 \times 9$ sudokus, for example.

  In some problems, one is interested in regularities that  often but not always occur.
  An example would be regularities in certain types of buildings.
  While certain regularities would occur in many buildings, there are often exceptions and different styles.
  In this case, one could attempt to find constraints on the structure of the floor plans that are satisfied by a given amount of examples.
\end{example}

\begin{framed}
  \noindent
  \begin{minipage}{\textwidth}
    \paragraph*{Goal 2}
    Given a set of examples and a set of partial rankings over these examples, the goal of the optimization system is to learn soft constraints and weights such that the order described by these optimization criteria maximally corresponds with the given rankings.
  \end{minipage}
\end{framed}

The soft constraints consist of clauses that are satisfied by some of the examples.
Rankings are partial orders over the examples: $e_x = e_y > ... > e_a = e_b$.
The weighted clauses can be used to calculate a score for every example
\begin{eqnarray*}
  \sum\limits_{\mathit{weight}, \mathit{c}} \mathit{weight} \cdot v_{c}
\end{eqnarray*}
With $v_c = 1$ if the clause $c$ is satisfied by the example and $v_c = 0$ if $c$ is not satisfied.
These scores allow for a total order over all examples to be defined.

\begin{example}
  Consider the scenario of moving to a new city.
  Multiple job offers are available, different accommodations are to be considered and there are different schools in the city.
  Learning weighted soft constraints can make user preferences explicit.
  For example, it could be desirable that the school is in a low crime area or that the locations of the school and work place are close.
\end{example}

A third goal of this research is to show how these constraints and especially the optimization criteria can be used in practice to find an optimal solution.

\section{Approach}
\begin{figure}

  \centering
    \includegraphics[width=0.8\linewidth]{ApproachOverview.pdf}
  \caption{Overview approach}
  \label{fig:struktuur}

\end{figure}

In this research, both constraint learning and learning optimization criteria were implemented.
An overview is presented in figure~\ref{fig:struktuur}.
Clauses are learned using examples.
To find weighted clauses, first the constraint learning system is used to find soft constraints.
Then the weights for these soft constraints are determined based on the rankings.
In order to determine the weights, the \svm system is used.

\subsection{Input}
The input consists of examples and definitions.
Every example contains a set of constants that are part of the model and an exhaustive list of the relations that hold on these constants.
Definitions describe either typing information or predicate definitions.
Typing information relates directly to the problem domain, improves the accuracy and efficiency of the learned clauses, and is usually easy to provide for the user.
Predicate definitions describe the types of the terms occurring in a predicate.
Two special types of predicates are allowed: calculated predicates and symmetric predicates.
Calculated predicates are not listed in examples, instead background knowledge is used to generate them.
Symmetric predicates are predicates for which the order of the terms does not matter.
If optimization criteria are to be learned, a set of rankings over the given examples is also provided.

\begin{example}
  Imagine a problem that describes humans with a type \textit{human} and predicates human(\textit{human}), male(\textit{human}) and female(\textit{human}).
  An example could contain the constants S and A, both of type \textit{human}, and the relations: human(S), human(A), male(S), female(A).
  Clauses to be learned would be: human(x) $\leftarrow$ male(x), human(x) $\leftarrow$ female(x),  female(x) $\lor$ male(x) $\leftarrow$ human(x) and \textit{false} $\leftarrow$ female(x) $\land$ male(x).
\end{example}
\subsection{Clausal Discovery}
The clause learning system is based on the clausal discovery algorithm.
The logical system IDP is used for two essential functions: $\mathit{covers}$ and $\mathit{entails}$.
The $\mathit{covers}$ function calculates which examples are covered by a clause.
The $\mathit{entails}$ function calculates whether an accepted clause is logically entailed by background knowledge or clauses that are already in the result set.

Every clause is subjected to two additional tests that attempt to eliminate redundant clauses early on.
The subset-test determines whether the clause is a superset of another clause that has been accepted earlier.
If such a clause exists and that clause covers the same examples as the new clause, the new clause is discarded.
Otherwise it will be tested whether the clause covers enough examples.
Clauses that do not cover enough examples are refined by the refinement operator.
The refined clauses will be added to the working queue.
If a clause covers enough examples, the $\mathit{entails}$ functions is calculated to decide whether the accepted clause should be added to the result set.

Refinement is an important step.
Clauses are extended to be more general and potentially cover more examples.
The refinement operator uses a list of atoms that can potentially be added to a clause.
This list is calculated in advance, given a maximal amount of variables.
Within the body and head of a clause, atoms may only be added in a specific order.
Additionally, several restrictions are in place.
Clauses must be connected, i.e. new atoms must always contain a variable that has already occurred.
Secondly, clauses must be range-restricted, which means that no new variables may be introduced in the head of the clause.
By looking at the positions of the atoms in the list, clauses can be reduced to a unique form.
This reduces the amount of duplicate clauses generated.

\subsection{Optimization}
The first step in finding weighted soft constraints is to identify soft constraints by using the constraint learning system and a low threshold.
Examples are characterized by the soft constraints (clauses) which it satisfies.
Therefore, every example can be translated to a vector of boolean features.
Each feature $f_i$ corresponds with one of the clauses and is \emph{1} if the clause covers the example and \emph{0} if it does not.
Existing machine learning software can be used to learn a linear \emph{scoring} function $\sum_i w_i \cdot f_i$ over these features based on the given rankings.
Hereby, weights are assigned to every feature to express its importance.
For an unseen example, the feature vector is computed by calculating what clauses cover the new example.
The scoring function can then be used to calculate a score for that example.

\begin{example}
  Consider examples~$e_1, e_2, e_3$, rankings $e_1 > e_2, e_2 > e_3$ and clauses $c_1, c_2, c_3$.
  If the clauses cover examples according to table~\ref{tbl:cover_examples}, then the function $(1 \cdot c_1) + (0\cdot c_2) + (-2\cdot c_3)$ perfectly models the given rankings.
  An example $e_4$ that is covered by~$c_1$ and~$c_2$, is then assigned a score of $1$.
  This score has no value in the absolute sense, it can only be used to compare it to other examples ranked by the same function.
  In this example clause~$c_1$ represents a desirable property, clause~$c_2$ is ignored because it does not influence the ranking and clause~$c_3$ represents an undesirable property.

  \begin{table}[!htp]
  \caption{Clause coverage}
  \label{tbl:cover_examples}
  \begin{tabularx}{\textwidth}{c|ccc|X}
    \textbf{Clauses}  &$c_1$    & $c_2$   & $c_3$   & \textbf{Feature vector}\\
    \toprule
    $e_1$         & Covers  &       &       & 1, 0, 0\\
    $e_2$         &       & Covers  &       & 0, 1, 0\\
    $e_3$         & Covers  & Covers  & Covers  & 1, 1, 1\\
  \end{tabularx}
  \end{table}

  Considering an extended form of the human example, the optimization function could be:
  1 $\cdot$ (funny(x)~$\leftarrow$~human(x)) + (-2) $\cdot$ (lazy(x)~$\leftarrow$~human(x)), for example.
  Clauses with a weight of zero are ignored.
\end{example}

The \svm{} software was chosen to find the scoring function since it is a robust and efficient system.
Contrary to many other ranking systems, it uses a linear model.
The weights that it assigns to the features can be used directly as weights for the optimization criteria.
The input format that is used by \svm{} is also supported by many other learn-to-rank implementations.
Therefore, other linear ranking systems such as Coordinate Ascent \cite{metzler2007linear} could also be plugged into the implementation if so desired.

@@

This optimization task can be solved in IDP, using inductive definitions, aggregates and minimization.
The only limitation is that the weights must be integer values.
Therefore, the weights of the clauses are divided by the smallest absolute weight, multiplied by a constant and rounded to the closest integer if necessary.
Removing this restriction for IDP is currently being researched and future releases might be able to deal with rational numbers directly.

In order to model the optimization problem in IDP, every clause $c_i$ with variables $v_1, ..., v_n$ is represented by a number $i$. For every clause a predicate $t(i)$ is added to capture the truth value of the clause.
A function $\mathit{cost}(i)$ specifies the cost of not satisfying the clause, which is equal to the integer weight of the clause.
\begin{eqnarray*}
  t(i) \Leftrightarrow \forall v_1, ..., v_n : c_i. \\
  cost(i) = w_i.
\end{eqnarray*}

Using $t$ and $\mathit{cost}$, a function $\mathit{actual}(i)$ is then defined in IDP as 0 if $t(i)$ and $\mathit{cost}(i)$ if $\lnot t(i)$.
This function is used in the term $\sum_i actual(i)$ to be minimized, which will allow IDP to search for an optimal solution.

\section{Evaluation}
Several experiments aim to measure the accuracy and efficiency of the learning systems for constraints and optimization criteria.

\subsection{Constraints}
Four problems have been used to evaluate constraint learning.
Two are existing problems: map coloring and sudoku.
Map coloring is a problem where countries are assigned colors and neighboring countries may not have the same color.
The examples to learn from each consist of a set of three countries that are correctly colored.
For sudoku a $4 \times 4$ solved sudoku is given.
Furthermore, there is the elevator problem with a soft constraint that holds on 2 of the 3 given examples.
The last problem is the co-housing problem with 4 hard constraints and 5 given examples.

\paragraph{Accuracy}
The essential constraints were found for all problems.
Additionally, constraints were found that describe some structure in the problems, such as that countries are never their own neighbor.
These kind of constraints may help a constraint solver to work more efficiently.

If the limits on the number of variables or literals per clause are too large, constraints are found that over-fit the training data.
These constraints are too specific and will exclude valid solutions.
On the other hand, if the chosen limits are too small, the necessary constraints will not be found.

\begin{table}
  \caption{Execution times overview (average over 8 runs)}
  \begin{tabularx}{\linewidth}{rl|ll}

@@

\paragraph{Efficiency}
Table~\ref{tbl:uitvoering} shows the execution time for several experiments, demonstrating the effect of the syntactical restrictions.
Moreover, all efficiency measures (i.e. symmetric predicates, range restriction and connected clauses) are able to improve the execution time.

Further experiments show that increasing the number of variables or literals per clause impacts the efficiency.
Especially the combination of more variables \emph{and} literals can steeply increase the execution time.
Therefore it would be useful to adapt these parameters dynamically.
Adding additional examples only increases the execution time by a constant factor.
Since adding more examples can help improve the accuracy, this trade-off is often worthwhile.

>>>>

\def\year{2015}
\documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{etoolbox}
\usepackage{color}
\usepackage{changepage}
\usepackage{csquotes}
\usepackage{tabularx,booktabs,multirow}
\usepackage{longtable}
\usepackage{framed}

@@

\title{Learning Constraints and Optimization Criteria}
\author{
  Samuel Kolb\\
  KU Leuven, Leuven, Belgium\\
  samuel.kolb@cs.kuleuven.be
}

@@

\begin{abstract}
This paper presents a system that is able to automatically learn both constraints from examples and optimization criteria from rankings.
Both hard and soft constraints can be learned and constraints are represented as first order logical clauses.
Choosing such clauses as representations enables the use of ILP techniques such as clausal discovery.
In this research, optimization criteria are represented as weighted clauses.
For all examples, the clause learning system could find the essential constraints efficiently.
The learning system for optimization criteria is shown to achieve accurate results and can learn criteria that identify the correct optimal solution even for few examples and noisy rankings.
Finally, a procedure is outlined that allows weighted first order logic clauses to be used for optimization in practice.
\end{abstract}

% -------------------------------------------------
% Introduction
% -------------------------------------------------

\section{Introduction}
This paper introduces an approach for learning constraints and optimization criteria.
Using constraints and optimization criteria such as the ones proposed in this paper allows problems to be modeled \emph{declaratively}.
This means that the representation is decoupled from the process of finding a (optimal) solution for the problem.
However, non-experts often experience difficulties when modeling a problem using constraints \cite{Wallace:PrinciplesCP}.
The first goal of this research is to automate the step of modeling a problem, in order to, amongst others, make constraint solving more accessible.
In this approach a few examples of what a user considers valid solutions are used by a computer system to find constraints.
These constraints can then be used by an existing constraint solver.

Constraint learning attempts to find constraints automatically given examples.
By using first-order logical clauses as representation for constraints, techniques from Inductive Logic Programming (ILP) can be used for learning.
Hereby, this research follows the same approach as \cite{Lallouet:LearningCP}, exploiting parallels between ILP and constraint learning.
In this case, the clausal discovery algorithm \cite{DeRaedt:ClausalDiscovery} is used to find constraints.

There has been increasing interest in formulating declarative models for constraint satisfaction as well as optimization problems.
By identifying soft constraints that are only satisfied by some examples and assigning weights to them based on their importance, an optimal solution can be sought.
The use of weighted soft constraints to model user preferences is inspired, amongst others, by the research of \cite{campigotto2011active}.
Their work examines attempts to obtain similar results in an interactive context for propositional constraints.
In this research, first-order constraints are used and optimization criteria are modeled using a first-order version of weighted MAX-SAT.
This paper also outlines how these constraints can be used in an existing solver to obtain an optimal solution in practice.

Since the underlying goal of this research is to facilitate the generation of a formal representation, the input format aims to be easy to use.
It has been a key focus to make reasonable assumptions concerning the knowledge a user can provide.
One important difference compared to similar research (including \cite{campigotto2011active}) is that users provide partial rankings over examples instead of absolute scores.

This paper is based on research that has been conducted during the course of my masters thesis~\cite{kolb2015thesis}.
The key contribution presented in this paper is a novel approach for learning weighted first order clauses from rankings to capture user preferences.
In this context an implementation of clausal discovery has been realized that supports soft constraints.
Moreover, an approach to obtaining optimal solutions for the weighted first order clauses is presented.

% -------------------------------------------------
% Background
% -------------------------------------------------

\section{Background}
In this section a brief overview of the important concepts and relevant literature will be provided.

\paragraph{Clauses}
First-order logical clauses consist of a disjunction of literals.
Literals are defined as being either an atom (positive literal) or a negated atom (negative literal).
Clauses are often expressed using a body and a head: $\mathit{head} \leftarrow \mathit{body}$.
Hereby, positive literals are grouped in the head and negative literals are grouped in the body.
%For example, $\lnot a_1 \lor \lnot a_2 \lor a_3 \lor a_4$ corresponds to the clause $a_3, a_4 \leftarrow a_1, a_2$.

Terms are used to reason about objects and consist of either constants denoting specific objects or variables that can represent any object.
In this paper constants will be written in uppercase ($\mathit{John}$) and variables in lowercase ($\mathit{x}$).
Relations between logical objects are represented by predicates, consisting of a predicate symbol and a set of terms (e.g. $\mathit{married(John, x)}$).
Predicates are atoms and can be assigned a truth value.

\paragraph{Learning constraints}
Constraint learning is a hard problem.
The search space of constraints is typically very large and, unlike in typical machine learning problems, there are usually only few examples to learn from.

@@

\paragraph{Clausal Discovery}
The clausal discovery algorithm proposed by \cite{DeRaedt:ClausalDiscovery} attempts to learn first order logic clauses from positive examples.
Using a refinement operator, the search space is traversed by incrementally generalizing clauses until they cover all examples, starting from the empty clause~($\square$).
The result set consists of the most specific clauses covering all examples.

\paragraph{Logic Solver}
The logic system IDP \cite{de2013prototype,wittocx2008idp} is used to calculate whether examples satisfy clauses and is used as a theorem prover.
It supports an extension of first order logic ($FO(\cdot)$) and can also be used to generate new solutions.
Background knowledge is given as an IDP theory.

\paragraph{Optimization}
The Boolean Satisfiability Problem (SAT) attempts to determine whether a propositional formula can be satisfied.
Formulas can be rewritten into a special form called the conjunctive normal form (CNF), which consists of a conjunction of clauses (disjunctions).
An extension of this problem is the Maximum Satisfiability Problem (MAX-SAT).
In MAX-SAT the amount of satisfied clauses is maximized.
By assigning positive integer weights to clauses and maximizing the sum of the weights of satisfied clauses, one obtains the weighted MAX-SAT problem.
The work by \cite{campigotto2011active} attempts to learn propositional formulas and weights automatically using examples and absolute scores.
Inspired by their results, this research tries to learn first-order logic clauses and weights using examples and rankings.
The resulting optimization problem can be seen as an extension of weighted MAX-SAT in which the clauses are in first-order logic.

% -------------------------------------------------
% Problem Statement
% -------------------------------------------------

\section{Problem Statement}
The research presented in this paper aims to learn constraints and optimization criteria based on positive examples and user preferences.
Since it aims to simplify the process of generating formal representations it should be easy to use and require input that a user can reasonably be expected to provide.

\subsection{Constraint learning}
\emph{Given a set of examples and a limit $t$, the goal is to find maximally specific clauses that are satisfied by at least $t$ of the given examples.}
\\\\
Examples consist of a domain and an exhaustive description of all relations that hold on the domain objects.
The learned clauses are domain-independent and contain only variables.

% \begin{example} " TODO - necessary?"
%   Consider sudoku as an example of a problem where hard constraints are to be found.
%   Given one or more solved sudokus, the learning system has to identify the rules of the game.
%   Since the learned clauses only contain universally quantified variables, for example $4 \times 4$ sudokus could be used to find constraints.
%   The learned constraints are independent of the size of the sudoku and can also be used to solve $9 \times 9$ sudokus, for example.

%   In some problems, one is interested in regularities that  often but not always occur.
%   An example would be regularities in certain types of buildings.
%   While certain regularities would occur in many buildings, there are often exceptions and different styles.
%   In this case, one could attempt to find constraints on the structure of the floor plans that are satisfied by a given amount of examples.
% \end{example}

\subsection{Learning optimization criteria}
\emph{Given a set of examples and a set of partial rankings over these examples, the goal is to identify soft constraints and weights learn such that the order described by these optimization criteria maximally corresponds with the given rankings.}
\\\\
The soft constraints consist of clauses that are satisfied by some but not all of the examples.
They can be identified using a system that implements the constraint learning task.
Partial rankings reason over sets of examples and describe an order over them (e.g. the ranking $e_i > e_j$ expresses the users preference of example $e_i$ over example $e_j$).
The weighted clauses (i.e. tuples $(weight, clause)$) can be used to calculate a score for every example:
\begin{eqnarray*}
  score(e) = \sum\limits_{(\mathit{w}, \mathit{c})} \mathit{w} \cdot \mathbf{1}_{c \text{ covers } e}
\end{eqnarray*}

\begin{example}[Moving]
  \label{ex:moving}
  Consider the scenario moving to a new city.
  The user has to choose a job, a place to live and a school.
  There are multiple job offers, houses and schools available across different areas of the city.
  User provided examples include a choice of where to work, where to live and what school is chosen.
  By comparing and ranking some of the examples, the user can express his preferences.
  Weighted soft-constraints can now be learned and used to identify the best choice.
  In this case, a possible soft constraint could be that the school and work place are in the same area.
  The weights that have been learned indicate how desirable or how undesirable a constraint is.
  Since the constraints are domain independent they can be used for any (unseen) city.
\end{example}

Aside from learning optimization criteria, this paper presents an approach to use these optimization criteria to find an optimal solution in practice.

% -------------------------------------------------
% Approach
% -------------------------------------------------

\section{Approach}
In this research, both constraint learning and learning optimization criteria were implemented.
Clauses are learned using examples.
The clause learning implementation uses the IDP system to test if clauses satisfy the examples and identify redundant clauses.
When learning weighted clauses, soft constraints are first identified in the examples using the clause learning implementation.
These constraints are assigned weights based on the given rankings.

\subsection{Input}
The input consists of global definitions and a set of examples.
Definitions consist of either typing information or predicate definitions.
Typing information relates directly to the problem domain, improves the accuracy and efficiency of the learned clauses and is usually easy to provide for the user.
Predicate definitions define predicate symbols and the types of their terms.
Two special types of predicates are allowed: calculated predicates and symmetric predicates.
Calculated predicates are not included in examples, instead background knowledge is used to generate them.
This background knowledge is passed directly to the IDP system.
Symmetric predicates are predicates for which the order of the terms does not matter.
Every example contains a domain consisting of a set of constants and an exhaustive list of the relations (predicates) that hold on these constants.
If optimization criteria are to be learned, a set of pairwise over the given examples is also provided.
Rankings are used instead of absolute scores since rankings only concern few examples at a time.
Because rankings only ask for relative positioning of a few examples, they are easier to provide.

\begin{example}
  \label{ex:moving-input}
  Consider the moving problem (example~\ref{ex:moving-input}).
  A model could, for example, use a type \textit{Area} and predicate definitions low\_crime(\textit{Area}), cheap(\textit{Area}), school\_in(\textit{Area}), work\_in(\textit{Area}) and live\_in(\textit{Area}).
  Examples would that specify various areas $A_i$ and describe the conditions (e.g. cheap($A_1$), low\_crime($A_2$)) as well as the users choices (e.g. school\_in($A_2$), work\_in($A_2$), live\_in($A_1$)).
  % Imagine a problem that describes humans with a type \textit{human} and predicates human(\textit{human}), male(\textit{human}) and female(\textit{human}).
  % An example could contain the constants S and A, both of type \textit{human}, and the relations: human(S), human(A), male(S), female(A).
  % Clauses to be learned would be: human(x) $\leftarrow$ male(x), human(x) $\leftarrow$ female(x),  female(x) $\lor$ male(x) $\leftarrow$ human(x) and \textit{false} $\leftarrow$ female(x) $\land$ male(x).
\end{example}

\begin{algorithm}
  \caption{The clausal discovery algorithm}
  \label{alg:cd}

  \begin{algorithmic}
  \State $Q \gets \{\square\}$, $\sym{T} \gets \{\}$
  \While{$\lnot isempty(Q)$}
    \State $c \gets next(Q)$
    \If{$covers(c, \textit{Examples})$}
      \If{$\lnot entails(\sym{T}, c)$}
        \State $\sym{T} = \sym{T} \cup c$
      \EndIf
    \Else
      \State $Q \gets Q \cup \rho(c)$
    \EndIf
  \EndWhile
  \State $\sym{T} \gets prune(\sym{T})$
  \State \Return \sym{T}
  \end{algorithmic}
\end{algorithm}

\subsection{Clausal Discovery}
The clause learning system is based on the clausal discovery algorithm (alg.~\ref{alg:cd}).
The logical system IDP is used for two essential functions: $\mathit{covers}$ and $\mathit{entails}$.
The $\mathit{covers}$ function calculates which examples are covered by a clause.
The $\mathit{entails}$ function calculates whether an accepted clause is logically entailed by background knowledge or clauses that are already in the result set.

Starting with the empty clause, clauses are refined until they cover enough examples.
If a clause covers enough examples it is added to result set, provided it is not entailed by background knowledge or other clauses in the result set.
To avoid extensive calls to $\mathit{covers}$ and $\mathit{entails}$, two additional tests are used.
The subset-test rejects clauses that are supersets of a clause that covers the same or all examples and has been accepted earlier.
Because clauses can be formulated in multiple ways, a representative-test is used to remove clauses that are not in canonical form.

In the refinement step, clauses are extended to be more general and potentially cover more examples.
The refinement operator uses a list of atoms that may be added to a clause.
This list is calculated in advance, given a maximal amount of variables.
Within the body and head of a clause, atoms may only be added in a specific order to avoid generating redundant clauses.
Object Identity is used to specify that variables with different names must denote different objects.
Additionally there are two syntactic restrictions.
Clauses must be connected (i.e. new atoms must always contain a variable that has already occurred) and range-restricted (i.e. no new variables may be introduced in the head of the clause).

\subsection{Optimization}
The first step in finding weighted soft constraints is to identify soft constraints by using the constraint learning system with a (low) threshold.
Examples are characterized by the soft constraints (clauses) which they satisfy.
Therefore, every example~$e$ can be translated to a vector of boolean features by introducing a feature~$f_i$ for every clause~$c_i$, with $f_i = \mathbf{1}_{c_i \text{ covers } e}$.
Existing software can be used to learn a linear scoring function $\sum_i w_i \cdot f_i$ over these features based on the given rankings.
For an unseen example, the feature vector is computed by calculating what clauses cover the new example.
The scoring function can then be used to calculate a score for that example.

\begin{example}
  Consider examples~$e_1, e_2, e_3$, rankings $e_1 > e_2, e_2 > e_3$ and clauses $c_1, c_2, c_3$.
  If the clauses cover examples according to table~\ref{tbl:cover_examples}, then, for example, the function $(1 \cdot c_1) + (0\cdot c_2) + (-2\cdot c_3)$ perfectly models the given rankings.
  A new example that is covered by~$c_1$ and~$c_2$, would be assigned a score of $1$, according to this function.
  This score has no value in the absolute sense, it can only be used to compare it to other examples ranked by the same function.
  In this example clause~$c_1$ represents a desirable property, clause~$c_2$ is ignored because it does not influence the ranking and clause~$c_3$ represents an undesirable property.

  \begin{table}
  \caption{Clause coverage}
  \label{tbl:cover_examples}
  \begin{tabularx}{\linewidth}{c|l|X}
    \textbf{Example} & \textbf{Covered by} & \textbf{Feature vector}\\
    \toprule
    $e_1$             & $c_1$               & (1, 0, 0) \\
    $e_2$             & $c_2$               & (0, 1, 0) \\
    $e_3$             & $c_1$, $c_2$, $c_3$ & (1, 1, 1) \\
  \end{tabularx}
  \end{table}
\end{example}

The \svm{} software was chosen to find the scoring function since it uses a linear model and offers an efficient implementation.
The weights that it assigns to the features can be used directly as weights for the optimization criteria.
The input format that is used by \svm{} is also supported by many other learn-to-rank implementations.
Therefore, other linear ranking systems such as Coordinate Ascent \cite{metzler2007linear} could also be used.

@@

This optimization task can be solved in IDP, using inductive definitions, aggregates and minimization.
The only limitation is that the current version only supports integer values.
Therefore, the weights of the clauses are divided by the smallest absolute weight, multiplied by a constant and rounded to the closest integer if necessary.

In order to model the optimization problem in IDP, every clause $c_i$ with variables $v_1, ..., v_n$ is represented by a number $i$. For every clause a predicate $t(i)$ is added to capture the truth value of the clause.
A function $\mathit{cost}(i)$ specifies the cost of not satisfying the clause, which is equal to the weight of the clause.
\begin{eqnarray*}
  t(i) \Leftrightarrow \forall v_1, ..., v_n : c_i. \\
  cost(i) = w_i.
\end{eqnarray*}

Using $t$ and $\mathit{cost}$, a function $\mathit{actual}(i)$ is then defined in IDP as $\mathbf{1}_{\lnot t(i)} \cdot \mathit{cost}(i)$.
This function is used in the term $\sum_i actual(i)$ to be minimized, which will allow IDP to search for an optimal solution.

% -------------------------------------------------
% Evaluation
% -------------------------------------------------

\section{Evaluation}
Several experiments aim to measure the accuracy and efficiency of the learning systems for constraints and optimization criteria.
In order to account for non-determinism, experiments measuring execution times or optimization scores are usually performed eight times and the results are averaged.

\subsection{Constraints}
Four problems have been used to evaluate constraint learning.
The first problem is map coloring, where countries are assigned colors and neighboring countries may not have the same color.
Two examples containing each three correctly colored countries are given.
The second problem is sudoku and a single, solved $4 \times 4$ sudoku is given.
For the third problem (elevator) three examples have been generated, two of which respect the underlying soft constraint.
The last problem (co-housing) contains four hard constraints and five examples have been generated that respect the constraints.

\begin{question}[Accuracy]
  Are the essential constraints discovered and what influence do different parameters have on the accuracy?
\end{question}
For all problems the essential constraints are found.
Often additional constraints were found that describe some structure in the problems.
For example, for the map coloring problem a learned constraint states that countries are never their own neighbor.
These kind of constraints may help a constraint solver to work more efficiently.

The learning process is parameterized by the maximal amount of variables and literals allowed per clause.
If these limits are too large, constraints are found that over-fit the training data.
These constraints are too specific and will exclude valid solutions that are not in the training set.
On the other hand, if the chosen limits are too small, the necessary constraints will not be found.

Over-fitting can be addressed by removing constraints that exclude valid solutions (manually or automatically) as well as providing more (or larger) training examples.
Negative examples can be used to detect under-fitting, indicating that the parameter values are too small.

\begin{table}
  \caption{Execution times overview}
  \begin{tabularx}{\linewidth}{rl|ll}

@@

\begin{question}[Efficiency]
  How fast is the clause learning system and what is the effect of various design decisions and input on the execution time?
\end{question}

Table~\ref{tbl:uitvoering} shows the execution times for several experiments.
It shows that smaller problems can be solved efficiently and shows that removing the syntactical restrictions exponentially increase the search space.

All efficiency measures (i.e. symmetric predicates, subset test and representative test) have been able to improve the execution time, sometimes by more than 50\%.
The overhead they introduce is more than compensated by the efficiency gains they cause.

Further experiments show that increasing the number of variables or literals per clause impacts the efficiency.
Especially the combination of more variables \emph{and} literals can steeply increase the execution time.
Therefore it would be useful to adapt these parameters dynamically.

Adding additional examples only increases the execution time by a constant factor.
Since adding more examples can help improve the accuracy, this trade-off is often worthwhile.
